The evolution of gene expression in the term placenta of viviparous mammals

### 2016-02-22 presentation/placenta_viviparous_presentation.Rnw
update presentation

\documentclass[ignorenonframetext]{beamer}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
% \usepackage{multirow}
\usepackage{acronym}
\usepackage{qrcode}
\usepackage{cleveref}
\usepackage[absolute,overlay]{textpos}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
\addbibresource{references.bib}

@@

\mode<presentation>{ 
  \usetheme{CambridgeUS}
  \usecolortheme{crane}
%   % http://identitystandards.illinois.edu/graphicstandardsmanual/generalguidelines/colors.html
  \definecolor{ilboldblue}{HTML}{002058}
  \definecolor{ilboldorange}{HTML}{E87722}
  \definecolor{ilblue}{HTML}{606EB2}
  \definecolor{ilorange}{HTML}{D45D00}
%   \setbeamercolor{alerted text}{fg=ilboldorange}
%   \setbeamercolor*{palette primary}{fg=ilboldblue,bg=ilorange!70!white}
%   \setbeamercolor*{palette secondary}{fg=ilboldblue,bg=ilorange!60!white}
%   \setbeamercolor*{palette tertiary}{bg=ilblue!60!white,fg=ilboldorange}
%   \setbeamercolor*{palette quaternary}{fg=ilboldblue,bg=ilorange!80!white}
%   
%   \setbeamercolor*{sidebar}{fg=ilorange,bg=ilboldblue}
%   
%   \setbeamercolor*{palette sidebar primary}{fg=ilblue!10!white,bg=ilorange}
%   \setbeamercolor*{palette sidebar secondary}{fg=ilorange}
%   \setbeamercolor*{palette sidebar tertiary}{fg=ilblue}
%   \setbeamercolor*{palette sidebar quaternary}{fg=ilorange}
%   
%   % \setbeamercolor*{titlelike}{parent=palette primary}
%   \setbeamercolor{titlelike}{parent=palette primary,fg=ilboldblue,bg=ilorange}
%   \setbeamercolor{frametitle}{fg=ilboldorange,bg=ilblue!20!white}
%   \setbeamercolor{frametitle right}{fg=ilboldblue,bg=ilorange}
%   
%   \setbeamercolor*{separation line}{}
%   \setbeamercolor*{fine separation line}{}
%   \setbeamercovered{transparent}  
  \logo{\begin{tikzpicture}% Pale figure
      {\node[opacity=0.734]{\IfFileExists{figures/igb_wordmark.pdf}{\includegraphics[width=2cm,height=1cm,keepaspectratio]{figures/igb_wordmark}}{}%
        };}%
    \end{tikzpicture}}
}

@@

\begin{frame}{Variation in Gestational Periods and Litter Size}
\tiny
\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
%  \caption{Gestational periods in days and average liter sizes for studied species.}
%  \label{tab:gestational_periods}
\end{table}
\end{frame}

\begin{frame}{Variations in Placenta Morphology}
\begin{table}
\begin{tabular}{c c c}
  \toprule
  Species & Interhemal Membrane & Pattern \\
  \midrule
  \smonkey & hemochorial & trabecular \\
  \cow     & epitheliochorial & villous \\
  \dog     & endotheliochorial & labyrinthine \\
  \armadillo & hemochorial & trabecular \\
  \horse & epitheliochorial & villous \\
  \elephant & endotheliochorial & labyrinthine \\
  \human & hemochorial & villous \\
% http://onlinelibrary.wiley.com/doi/10.1002/jez.1402660511/pdf
  \opossum & epitheliochorial & choriovitelline \\
  \mouse & hemotrichorial & labyrinthine \\
  \galili & hemotrichorial & labyrinthine \\
  \sheep & epitheliochorial & villous \\
  \bonobo & hemochorial & villous \\
% http://www.biolreprod.org/content/early/2014/10/17/biolreprod.114.124818.short
  \carmeli & hemotrichorial & labyrinthine \\
  \pig & epitheliochorial & diffuse \\
  \bottomrule
\end{tabular}
\end{table}
\end{frame}

@@

placenta.lineage.specific.table <- function(analysis,output=TRUE) {
    temp.table <- 
        head(placenta.lineage.specific.analysis[analysis.name==analysis &
                                                p.value <= 0.05][order(p.value),
                                                                 list(human_name,p.value,fdr,fc),],
             n=5)
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[order(`p (t-test)`),]
        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        print(xtable(display.table,
                     display=c("s","s","s","e","g","f"),
                     columns=c("c","c","p{1.5in}","c","c","c")
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

@@

\item We considered a gene to be core to the placenta transcriptome if
  a 1:1 ortholog of that gene was expressed with
  $\mathrm{FPKM} \ge 10$ in all species studied and was not a human
  housekeeping gene according to
  \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.

@@

\item ANXA1 is involved in the resolution of the inflammation response
  through inhibition of phospholipase A2 activity and signaling
  through the formyl peptide receptor family, and has previously been
  hypothesized to be important for the maintenance of the
  anti-inflammatory during pregnancy
  \cite{Hutchinson.ea2011:Molecularregulatorsofresolution}.

\item ANXA2 is known to interact with S100A10, S100A6, S100A11, and
  S100A4 \cite{Liu.ea2015:AnnexinA2complexeswith}, and there is
  evidence for its interaction with S100P as well; it is involved in
  cell-cell interaction as well as vesicle trafficking and von
  Willibrand Factor secretion.

\item The expression regulation of another core gene, \textit{TIMP3},
  has been implicated in pre-eclampsia
  \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
\end{itemize}

@@

\begin{frame}{PSG1}
\begin{figure}
<<psg.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
plot.label.size <- 3
axis.text.size <- 3.5

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
                 axis.text.size=axis.text.size,
                 fontsize=plot.label.size,
                 )
@ 
\end{figure}
\end{frame}

\begin{frame}{CTSL}
\begin{figure}
<<ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
plot.label.size <- 3
axis.text.size <- 3.5
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
                 fontsize=4,
                 width=3,
                 axis.text.size=axis.text.size,
                 offset.ratio=1
                 )

\begin{frame}{PAG2}
\begin{figure}
<<pag2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSSSCG00000006792",
                           "ENSBTAG00000046085"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
                 axis.text.size=axis.text.size,
                 fontsize=plot.label.size,
                 offset.ratio=0.6
                 )

@@

\begin{frame}{Galectin Expression}
\begin{figure}
<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=

@@

igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))

@@

common.plot.options <- 
    list(geom_tile(aes(fill = fpkm), colour = "white"),
##         scale_fill_gradientn(colours = colorends, values = rescale(gradientends)),
         scale_fill_gradient(low = grad.def[["low"]],
                             high = grad.def[["high"]],
                             space = grad.def[["space"]],
                             na.value = grad.def[["na.value"]],
                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
         scale_x_discrete("", expand = c(0, 0)),
         scale_y_discrete("", expand = c(0, 0)),
         theme_grey(base_size = 9),
         theme(legend.position = "top",
               axis.ticks = element_blank(),
               panel.background = element_blank(),
               axis.text.y = element_text(face="italic"),
               axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
         facet_grid(type~.,scale="free_y",space="free_y")
         )
print(ggplot(combined.long[combined.long$type=="Galectin",], aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\caption{Expression of Hormone Genes, Galectins, and IGF related genes}
\label{fig:hormone_genes}
\end{figure}
\end{frame}

\begin{frame}{Hormone Expression}
\begin{figure}
<<hormone.expression.plot,echo=FALSE>>=
print(ggplot(combined.long[combined.long$type=="Hormone",], aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\end{figure}
\end{frame}

\begin{frame}{IGF Related}
\begin{figure}
<<igfrelated.expression.plot,echo=FALSE>>=
print(ggplot(combined.long[combined.long$type=="IGF Related",], aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\end{figure}
\end{frame}

@@

\begin{frame}{Growth Hormone family}

\begin{columns}
\column{0.4\textwidth}
The \ac{GH} family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two).

@@

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,
                 axis.text.size=5,
                 fontsize=4)
@ 
\end{columns}

@@

  The human gene \textit{CGA} codes for the alpha subunit of multiple
  hormones including pituitary hormones (luteinizing hormone,
  follicle-stimulating hormone, thyroid-stimulating hormone), as well
  as numerous chorionic gonadotropins with variable beta subunits
  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
  the chorionic gonadotropins is to prevent the regression of the
  corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six
  placentas we analyzed here, \textit{CGA} is only expressed in the
  placentas of \human and \bonobo
  (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}), which
  corresponds with the emergence of the chorionic gonadotropins in
  anthropoid primates
  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
  \textit{CGA} is expressed at much higher levels in both species than
  the beta subunits with which its product dimerizes \textit{CGA} is
  one of several genes noted by
  \citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
  adaptive evolution of the \human{}-\textit{P. troglodytes} stem
  lineage.

\begin{figure}
<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

plot.label.size <- 4.5
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 offset.ratio=0.7,
                 fontsize=plot.label.size,
                 axis.text.size=3.5
                 )
@ 
\end{figure}

>>>>

\documentclass[ignorenonframetext]{beamer}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
\usepackage{pdfpcnotes}
% \usepackage{multirow}
\usepackage{acronym}
\usepackage{qrcode}
\usepackage{cleveref}
\usepackage[absolute,overlay]{textpos}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
\addbibresource{references.bib}

@@

\mode<presentation>{ 
  \usetheme{CambridgeUS}
  \usecolortheme{crane}
%   % http://identitystandards.illinois.edu/graphicstandardsmanual/generalguidelines/colors.html
  \definecolor{ilboldblue}{HTML}{002058}
  \definecolor{ilboldorange}{HTML}{E87722}
  \definecolor{ilblue}{HTML}{606EB2}
  \definecolor{ilorange}{HTML}{D45D00}
%   \setbeamercolor{alerted text}{fg=ilboldorange}
%   \setbeamercolor*{palette primary}{fg=ilboldblue,bg=ilorange!70!white}
%   \setbeamercolor*{palette secondary}{fg=ilboldblue,bg=ilorange!60!white}
%   \setbeamercolor*{palette tertiary}{bg=ilblue!60!white,fg=ilboldorange}
%   \setbeamercolor*{palette quaternary}{fg=ilboldblue,bg=ilorange!80!white}
%   
%   \setbeamercolor*{sidebar}{fg=ilorange,bg=ilboldblue}
%   
%   \setbeamercolor*{palette sidebar primary}{fg=ilblue!10!white,bg=ilorange}
%   \setbeamercolor*{palette sidebar secondary}{fg=ilorange}
%   \setbeamercolor*{palette sidebar tertiary}{fg=ilblue}
%   \setbeamercolor*{palette sidebar quaternary}{fg=ilorange}
%   
%   % \setbeamercolor*{titlelike}{parent=palette primary}
%   \setbeamercolor{titlelike}{parent=palette primary,fg=ilboldblue,bg=ilorange}
%   \setbeamercolor{frametitle}{fg=ilboldorange,bg=ilblue!20!white}
%   \setbeamercolor{frametitle right}{fg=ilboldblue,bg=ilorange}
%   
%   \setbeamercolor*{separation line}{}
%   \setbeamercolor*{fine separation line}{}
%   \setbeamercovered{transparent}  
  \logo{\begin{tikzpicture}% Pale figure
      {\node[opacity=0.5]{\IfFileExists{figures/igb_wordmark.pdf}{\includegraphics[width=2cm,height=1cm,keepaspectratio]{figures/igb_wordmark}}{}%
        };}%
    \end{tikzpicture}}
}

@@

\begin{frame}{Variation in Gestational Periods and Litter Size}
\footnotesize
%\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
%  \caption{Gestational periods in days and average liter sizes for studied species.}
%  \label{tab:gestational_periods}
%\end{table}
\end{frame}

\begin{frame}{Variations in Placenta Morphology}
%\begin{table}
\footnotesize
\begin{tabular}{c c c}
  \toprule
  Species & Interhemal Membrane & Pattern \\
  \midrule
  \smonkey & hemochorial & trabecular \\
  \cow     & epitheliochorial & villous \\
  \dog     & endotheliochorial & labyrinthine \\
  \armadillo & hemochorial & trabecular \\
  \horse & epitheliochorial & villous \\
  \elephant & endotheliochorial & labyrinthine \\
  \human & hemochorial & villous \\
% http://onlinelibrary.wiley.com/doi/10.1002/jez.1402660511/pdf
  \opossum & epitheliochorial & choriovitelline \\
  \mouse & hemotrichorial & labyrinthine \\
  \galili & hemotrichorial & labyrinthine \\
  \sheep & epitheliochorial & villous \\
  \bonobo & hemochorial & villous \\
% http://www.biolreprod.org/content/early/2014/10/17/biolreprod.114.124818.short
  \carmeli & hemotrichorial & labyrinthine \\
  \pig & epitheliochorial & diffuse \\
  \bottomrule
\end{tabular}
%\end{table}
\end{frame}

@@

placenta.lineage.specific.table <- function(analysis,output=TRUE) {
    temp.table <- 
        head(placenta.lineage.specific.analysis[analysis.name==analysis &
                                                p.value <= 0.05][order(p.value),
                                                                 list(human_name,p.value,fdr,fc),],
             n=5)
    temp.table <- data.table(temp.table)
    setkey(temp.table,"human_name")
    temp.table <- 
        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
                                            `Approved Name`,p.value,fdr,fc)]
    setnames(temp.table,
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
              )
    setkey(temp.table,"Human Symbol")
    if (output) {
        display.table <- temp.table[order(`p (t-test)`),]
        display.table <-
            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
                                                  "}")]
        print(xtable(display.table,
                     display=c("s","s","s","e","g","f"),
                     align=c("c","c","p{1.5in}","c","c","c")
                     ),
              booktabs=TRUE,
              floating=FALSE,
              math.style.exponents=TRUE,
              sanitize.text.function=genes.table.sanitize,
              include.rownames=FALSE)

@@

\item Gene is core to the placenta transcriptome if a 1:1 ortholog is
  expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
  not a human housekeeping gene according to
  \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.

@@

\item ANXA1 is involved in resolution of inflammation by inhibiting
  phospholipase A2 activity and signaling via formyl peptide receptor
  family
\item hypothesized to be important for the maintenance of the
  anti-inflammatory during pregnancy
  \cite{Hutchinson.ea2011:Molecularregulatorsofresolution}.

\item ANXA2 interacts with S100A10, S100A6, S100A11, and
  S100A4 \cite{Liu.ea2015:AnnexinA2complexeswith}, and there is
  evidence for its interaction with S100P as well
\item ANXA2 involved in cell-cell interaction as well as vesicle
  trafficking and von Willibrand Factor secretion.

\item \textit{TIMP3} has been implicated in pre-eclampsia
  \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
\end{itemize}

@@

\begin{frame}{Pregnancy-Specific Glycoprotein Expansion}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item 11 highly expressed \textit{PSG} genes in \human and some
  expression in \smonkey and \bonobo
\item \mouse specific expansion and high expression of \textit{PSG} as
  well
\item No other species have high expression of \textit{PSG} or
  \textit{CEACAM}
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<psg.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
plot.label.size <- 3
axis.text.size <- 3.5

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
                 axis.text.size=axis.text.size,
                 fontsize=plot.label.size,
                 )
@ 
\end{figure}
\end{columns}
\end{frame}

\begin{frame}{Cathepsin-L expansion}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item Cathepsins are highly expressed in many placentas
\item Expanded in \mouse and \galili and possibly also \carmeli to a
  lesser degree
\item Intracellular protease which degrades collagen, elastin, and in
  IGF1R
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
plot.label.size <- 3
axis.text.size <- 3.5
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
                 fontsize=4,
                 width=3,
                 axis.text.size=axis.text.size,
                 offset.ratio=1
                 )

@ 
\end{figure}
\end{columns}
\end{frame}

\begin{frame}{Pregnancy associated glycoprotein}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item Massive expansion in ruminants (\cow and \sheep)
\item 28 members of this family have $\mathrm{FPKM} \ge 1$ in \cow
\item \Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}
\item \Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<pag2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=1,
                 subtree=c("ENSSSCG00000006792",
                           "ENSBTAG00000046085"),
                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
                 axis.text.size=axis.text.size,
                 fontsize=plot.label.size,
                 offset.ratio=0.6
                 )

@ 
\end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}{Galectin Expression}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item Galectins are major regulators of pregnancy in fetal and
  maternal tissues \cite{Than.ea2012:Galectinsguardiansofeutherian}
\item \textit{LGALS1} expressed everywhere
\item Primate expansion of \textit{LGALS13} and \textit{LGALS14}
\item Likely present in ancestral placenta of therian mammals as it is
  expressed even in \opossum
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=

@@

igfrelated.genes <- unique(c(grep("^IGF",combined.fpkm[,human_name],value=TRUE),"H19"))

@@

common.plot.options <- 
    list(geom_tile(aes(fill = fpkm), colour = "white"),
##         scale_fill_gradientn(colours = colorends, values = rescale(gradientends)),
         scale_fill_gradient(low = grad.def[["low"]],
                             high = grad.def[["high"]],
                             space = grad.def[["space"]],
                             na.value = grad.def[["na.value"]],
                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
         scale_x_discrete("", expand = c(0, 0)),
         scale_y_discrete("", expand = c(0, 0)),
         theme_grey(base_size = 9),
         theme(legend.position = "right",
               axis.ticks = element_blank(),
               panel.background = element_blank(),
               axis.text.y = element_text(size=15,face="italic"),
               axis.text.x = element_text(size=16,angle = 300, hjust = 0, vjust=1,face="italic")),
         facet_grid(type~.,scale="free_y",space="free_y")
         )
print(ggplot(combined.long[combined.long$type=="Galectin",], aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\end{figure}
\end{columns}
\end{frame}

\begin{frame}{Hormone Expression}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item \human specific high expression of \textit{CGA}
\item \human expansion of \textit{CGB}
\item Vasodilation-involved proteins (\textit{NPPB} \& \textit{NPPC})
  highly-expressed in \elephant and \dog, but not expressed elsewhere
  \pnote{NPPB and NPPC have been seen in decidua of mouse
    \cite{Cameron.ea1996:sitesofgeneexpression}.}
\item \textit{PRL} highly expressed in \elephant
\pnote{more about PRL later}
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<hormone.expression.plot,echo=FALSE>>=
print(ggplot(combined.long[combined.long$type=="Hormone",], aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\end{figure}
\end{columns}
\end{frame}

\begin{frame}{IGF Related}
\begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item \textit{IGF2} is highly expressed in most clades
\item \textit{IGF2} promotes the growth and division of cells; highly active during fetal development
\item \textit{IGF2} binding partners are also highly expressed
  (\textit{IGFBP3}, \textit{IGFBP5})
\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<igfrelated.expression.plot,echo=FALSE>>=
print(ggplot(combined.long[combined.long$type=="IGF Related",], aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}{Growth Hormone (GH) family}

\begin{columns}
\column{0.4\textwidth}
The GH family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two).

@@

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,
                 offset=2,
                 axis.text.size=5,
                 fontsize=10)
@ 
\end{columns}

@@

  \begin{columns}
\column{0.4\textwidth}
\begin{itemize}
\item \textit{CGA} codes for the alpha subunit of multiple hormones as
  well as numerous chorionic gonadotropins with variable beta subunits
  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}.

\item Chorionic gonadotropins prevent the regression of the corpus
  luteum \cite{Hanson.ea1971:EffectsofHCGand}.

\item \textit{CGA} only expressed in the placentas of \human and
  \bonobo

  \pnote{Expression corresponds with emergence of the chorionic gonadotropins in
    anthropoid primates
    \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}.}

\end{itemize}
\column{0.6\textwidth}
\begin{figure}
<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

plot.label.size <- 4.5
tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 offset.ratio=1,
                 fontsize=8,
                 axis.text.size=5
                 )
@ 
\end{figure}
\end{columns}
\end{frame}
